---
title: "Week 3 - Lecture 1 - Probability"
format: html
editor: visual
author: 'Kierin Chung with scaffold and instructions from Yeeka Yau.'
---

## Chance simulation and Introduction to the Box model

```{r}
# This code suppresses some annoying warning messages
#| message: false
#| warning: false
#| 
library(tidyverse)
```

## Example: Two fair dice are thrown ðŸŽ² ðŸŽ². What is the chance that their sum is 6?

```{r}
# Set seed for reproducibility
set.seed(123)

# Create a "box" for the die
die <- c(1,2,3,4,5,6)

number_of_rolls = 1000

# Sample with replacement twice from the box to simulate rolling two fair die. Add up the numbers and repeat 1000 times
sums <- replicate(number_of_rolls,sum(sample(die,2,replace=TRUE)))

# Convert to data frame for ggplot
df <- data.frame(sample_sums = sums)

# Barplot of sample sums
ggplot(df, aes(x=sample_sums)) +
  geom_bar(fill = "skyblue", color = "white") +
  scale_x_continuous(breaks = 2:12) +
  labs(title="Sum of two dice rolls")
  theme_minimal()
```

### How many of the sums are equal to 6?

```{r}
table(sums)
```

### What is the probability that the sum will be greater than 9?

Based on the table, we have 96+49+24 sums greater than 9, so the probability is (96+49+24)/1000 $\approx$ 17% of the sum being greater than 9.

## Example: Basketball draft ðŸ€

```{r}
# Set seed for reproducibility
set.seed(3)

# Create a "box" for the player pool
basketball_box <- c(rep(1,8), rep(0,12))

number_of_drafts = 1000

# Sample WITHOUT replacement 5 times from the box to simulate a random selection of 5 players. Add up the numbers and repeat 1000 times
basketball_sums <- replicate(number_of_rolls,sum(sample(basketball_box,5,replace=FALSE)))

# Convert to data frame for ggplot
basketball_df <- data.frame(basketball_sum = basketball_sums)

# Barplot of sample sums
ggplot(basketball_df, aes(x=basketball_sum)) +
  geom_bar(fill = "purple", color = "white") +
  labs(title="Random draft of 5 players", x= "Number of star players")
  theme_minimal()
```

### What is the probability that exactly 2 star players are selected?

```{r}
table(basketball_sums)
```

Approximately 412/1000 = 0.412

### What is the probability that at least 1 star player is selected?

Approximately 1 - 50/1000 = 0.95

## Chevalier: Game A

```{r}
# Set seed for reproducibility
set.seed(4)

# Create a "box" for game A
game_A_box <- c(1,0,0,0,0,0)

number_of_games = 10000

# Sample WITHOUT replacement 5 times from the box to simulate a random selection of 5 players. Add up the numbers and repeat 10000 times
game_A_sums <- replicate(number_of_games,sum(sample(game_A_box,4,replace=TRUE)))

# Convert to data frame for ggplot
chevalier_df <- data.frame(game_A_sum = game_A_sums)

# Barplot of sample sums
ggplot(chevalier_df, aes(x=game_A_sum)) +
  geom_bar(fill = "orange", color = "white") +
  labs(title="Simulation of game A", x= "Number of aces")
  theme_minimal()
```

### Probability of getting at least 1 ace:

```{r}
table(game_A_sums)
```

Chance of winning: (3889+1146+159+10)/10000 $\approx$ 0.52

## Homework: Simulate Chevalier: Game B

```{r}
# Set seed for reproducibility
set.seed(62)

# Create a "box" for game A
game_B_box <- c(rep(0, 35), 1)

number_of_games = 10000

# Sample WITHOUT replacement 5 times from the box to simulate a random selection of 5 players. Add up the numbers and repeat 10000 times

game_B_sums <- replicate(number_of_games, sum(sample(game_B_box, 24, replace=TRUE)))

# Convert to data frame for ggplot
chevalier_df <- data.frame(game_B_sum = game_B_sums)

# Barplot of sample sums
ggplot(chevalier_df, aes(x=game_B_sum)) +
  geom_bar(fill = "orange", color = "white") +
  labs(title="Simulation of game B", x= "Number of aces")
  theme_minimal()
  
print(table(game_B_sums))
```

The probabiltiy he wins is ((10000-5117-3466)/10000) = 14.17%. This is because with the 36 different possibilities, it becomes harder to get 2 aces, therefore making it much more difficult to win. 
```